---
title: "Abundance Analysis"
author: "Kelly Endres"
date: "April 17, 2019"
output: html_document
---

In this code I am comparing visitor abundance between years. This is important because as is we are not able to look at straight abundances between years due to numerous sampling effort differences. We want to compare these as it is an important aspect of how drought may impact plant-pollinator communities. 

To examine abundance differences and account for sampling effort differences, we will be using a linear mixed effect model, glmmTMB adding several different fixed and random effects. 

First there are several specific packages necessary to complete the analysis.

```{r packages, include=FALSE}

#load packages
library(glmmTMB)
library(fitdistrplus)
library(DHARMa)
library(ggplot2)
library(tidyverse)
library(gridExtra)
library(viridis)
library(cowplot)

```

First step is to import the data needed for this code. I manually reformated the data in excel as it would take much less time and was more straightforward. These new data sheets have been uploaded separately into GitHub.

```{r data inport}

#inport data
ipo1997 <- read.csv("abundance_1997.csv") #1997
ipo1998 <- read.csv("abundance_1998.csv") #1998
ipo2001 <- read.csv("abundance_2001.csv") #2001
ipo2012 <- read.csv("abundance_2012.csv") #2012
ipo2018 <- read.csv("abundance_2018.csv") #2018

#'Year' in the 2012 dataset was actually entered as '2001'. Overwrote this
ipo2012$Year = 2012

```

With the data imported we need to combine all the years into one dataframe for analysis. 

```{r create dataframe}
#combine data
all.years <- rbind(ipo1997, ipo1998, ipo2001, ipo2012, ipo2018)

#rename site column
colnames(all.years)[1] <- "Site"

all.years$Year =as.factor(all.years$Year)
```

# Summary stats
```{r}
# Just averages across year-site 
all.years %>% 
  group_by(Year, Site) %>%
  tally(Richness)
  
all.years %>% 
  group_by(Drought) %>%
  summarize(mean_rich= mean(Richness))

# Mean richness was1.750 in drought years and 0.697 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_rich= mean(Richness))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_rich= mean(site_rich))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_rich))

# changes values slightly
# Non-drought	0.6931687			
# Drought	1.4957265	

# Abundance the same way
# Just averages across year-site 
all.years %>% 
  group_by(Year, Site) %>%
  tally(Visitor.Abundance)
  
all.years %>% 
  group_by(Drought) %>%
  summarize(mean_abund= mean(Visitor.Abundance))

# Mean abundance was 3.875 in drought years and 1.725 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_abud= mean(Visitor.Abundance))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_abund= mean(site_abud))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_abund))

# changes values slightly
# Non-drought	1.670445			
# Drought	3.183226	
```

The first step to examine visitor abundance between drought and non-drought years is to determine the correct distrubution for our data. It is most likely negative binomial, so we will start by checking for that. 

```{r abundance distribution fit}

#fit data distribution
distcheck <- fitdist(all.years$Visitor.Abundance, "nbinom")

#summarize nbinom fit
summary(distcheck)

#plot nbinom fit
plot(distcheck)
```

According the the plot of the distribution that data fits a negative binomial distribution. 

Now we are able to run our linear mixed effects model (glmmTMB) using a negative binomial distribution (nbinom2). Our responce variable is given as total visitor abundance for each inidividual site visit. Drought, observer number, and amount of time observing are fixed effects, while site nexted within year is a random effect. 

```{r Adbunace model}

abundancemod <- glmmTMB(Visitor.Abundance ~ Drought + Observers + Time + (1|Site/Year), all.years, family = "nbinom2")
sum= summary(abundancemod)

car::Anova(abundancemod)
aov(abundancemod)

tab_model(abundancemod, show.ci = FALSE)

summary(abundancemod)
# write.csv(sum.tab, "abundance.glmm.sum.csv") # not working
```

DESCRIPTION

```{r}
gg <- residuals(abundancemod)

plot(gg)

all.years$residuals <- gg

all.years$fitted <- abundancemod$fitted

simoutput <- simulateResiduals(fittedModel = abundancemod, n = 250)

plot(simoutput)
```

DESCRIPTION

```{r}
# create new column for variable
all.years$visithour <- all.years$Visitor.Abundance/all.years$Time
```

DESCRIPTION

```{r}
ggplot(all.years, aes(x=Drought, y=visithour)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(x="Drought", y="Visits per Hour") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title = element_text(hjust = 0.5, size= 14)) +
  scale_color_viridis(discrete=TRUE)


# change factor levels for color plotting
all.years$Drought <- factor(all.years$Drought, levels = c("Drought", "Non-drought"))

visit= ggplot(all.years, aes(x=Drought, y=Visitor.Abundance)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(colour = Drought), alpha = 0.5, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(y="Visitor Abundance", title = "(a)") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)

cow.a=ggplot(all.years, aes(x=Drought, y=Visitor.Abundance)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(colour = Drought), alpha = 0.5, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(y="Visitor Abundance") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)

```

Richness analysis with glmm


```{r}
c <- fitdist(all.years$Richness, "nbinom")

plot(c)

model3 <- glmmTMB(Richness ~ Drought + Observers + Time + (1|Site/Year), all.years, family = "nbinom2")

summary(model3)

# check residuals
gg2 <- residuals(model3)
plot(gg2)


all.years$residuals2 <- gg2

all.years$fitted2 <- model3$fitted

simoutput2 <- simulateResiduals(fittedModel = model3, n = 250)
plot(simoutput2)

richness= ggplot(all.years, aes(x=Drought, y=Richness)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(x="Drought", y="Richness", title = "(b)") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)

cow.b=ggplot(all.years, aes(x=Drought, y=Richness)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(x="Drought", y="Richness") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)


panel= grid.arrange(visit, richness, ncol=2)
ggsave("Figure 1 Abundance and Richness", panel, device="pdf", width= 16.6, height = 14.0, units= "cm")
```

Put plots 1a and 1b together
```{r}
# use cowplot
panel=plot_grid(cow.a, cow.b, labels = c('A', 'B'))

ggsave("Figure 1 Abundance and Richness_revision", panel, device="pdf", width= 16.6, height = 14.0, units= "cm")
```

