---
title: "Abundance Analysis"
author: "Kelly Endres"
date: "April 17, 2019"
output: html_document
---

In this code I am attempting to compare visitor and visit abundance between years. This is important because as is we are not able to look at differences in abundace due to numerous sampling effort differences. We want to compare these as it is an important aspect of how drought may impact plant-pollinator communities. 

To examine abundance differences and account for sampling effort differences, we will be using a linear mixed effect model, glmmADMB adding several different fixed and random effects. 


```{r setup, include=FALSE}
#clear environment
rm(list = ls())
```

The package that we want to use, glmmADMB, was for some reason taken down. However by running the following code it is possible to download it directly from the group that wrote it.

```{r packages, include=FALSE}

#install glmmADMB
install.packages("glmmADMB", 
     repos=c("http://glmmadmb.r-forge.r-project.org/repos", 
            getOption("repos")),
     type="source")

#load packages
library(glmmADMB)
```

First step is to import the data needed for this code. I manually reformated the data in excel as it would take much less time and we are in somewhat of a time crunch. 

```{r data inport}

#inport data
ipo1997 <- read.csv("abundance_1997.csv") #1997
ipo1998 <- read.csv("abundance_1998.csv") #1998
ipo2001 <- read.csv("abundance_2001.csv") #2001
ipo2012 <- read.csv("abundance_2012.csv") #2012
ipo2018 <- read.csv("abundance_2018.csv") #2018
```

With the data imported we need to combine all the years into one dataframe for analysis. 

```{r create dataframe}
#combine data
all.years <- rbind(ipo1997, ipo1998, ipo2001, ipo2012, ipo2018)

#rename site column
colnames(all.years)[1] <- "Site"
```



```{r run analysis}

a <- glmmadmb(Visitor.Abundance ~ Drought + Observers + Time + (Year|Site), all.years, family = "nbinom2", zeroInflation = TRUE)

summary(a)
```

