---
title: "Abundance Analysis"
author: "Kelly Endres"
date: "April 17, 2019"
output: html_document
---

In this code I am attempting to compare visitor and visit abundance between years. Due to numerous sampling effort differences this can be difficult. To account for this, we are converting visits into visits per person*hour (PH). After this, we will run mixed effect linear models. 


```{r setup, include=FALSE}
#clear environment
rm(list = ls())
```


```{r packages, include=FALSE}

#load packages

```

First step is to inport the data needed for this code. 

```{r data inport}

#inport data
ipo1997 <- read.csv("abundance_1997.csv") #1997
#ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("abundance_2001.csv") #2001
#ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("abundance_2018.csv") #2018
```

Because of some technical and time issues, I ended up reformating the data in excel. This ended up saving a lot of time!

Now that the reformated data has been inported, I am going to add all of the years into one dataframe. 


1997:

```{r}
#1997

#create list with time in hours
#Hours97 <- ipo1997$Time/60
#create list with person*time
#PH97 <- Hours97*ipo1997$Observers
#create visitor abundance per person*time
#VPH97 <- ipo1997$Visit/PH97
#create visit abundance per person*time
#TVPH97 <- ipo1997$Total.Visits/PH97

```

