---
title: "Adonis Analysis"
author: "Kelly Endres"
date: "January 22, 2019"
output: html_document
---


This set of analyses is to look at differences in community composition between drought and non-drought years. This requires construction  of community similarity/dissimilarity matrices and conducting Adonis analyses, with drought vs. nondrought as the fixed effects. Adonis is useful for this analysis as it uses strata which takes into account random effects (e.g. year) (???). We will use the Morisita-Horn index with Adonis as is is insensitve to differences in sampling effort among catagories. Therefore we can use raw data without correcting for sampling effort. 


```{r setup, include=FALSE}
#clear environment
rm(list = ls())
```


```{r packages, include=FALSE}

#load packages
library(vegan)
library(dplyr)
library(tidyverse)
library(reshape2)
```


```{r data inport}

#inport data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```


1997: 

```{r 1997 formating}

##1997

#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits (removing extra information)
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
tv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv97)[1] <- "1" #rename
tv97 <- tv97[, -1] #delete
#rename rows as sites
row.names(tv97) <- c("P97", "T97", "V97")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv97 <- cbind(tv97, "Other fly" = c(0, 0, 0), "Wasp" = c(0, 0, 0)) #add missing columns
tv97 <- tv97[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns alphabetically
#add year column
tv97 <- cbind(tv97, "Year" = rep("1997", length(tv97$Butterfly)))
#add drought column
tv97 <- cbind(tv97, "Drought" = rep("Non-drought", length(tv97$Butterfly)))
```

1998: 

```{r 1998 formatting}

##1998

#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
tv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv98)[1] <- "1" #rename
tv98 <- tv98[, -1] #delete
#rename rows as sites
row.names(tv98) <- c("P98", "T98", "V98")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv98 <- cbind(tv98, "Butterfly" = c(0, 0, 0), "Wasp" = c(0, 0, 0)) #add missing columns
tv98 <- tv98[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns alphabetically
#add year
tv98 <- cbind(tv98, "Year" = rep("1998", length(tv98$Butterfly)))
#add drought
tv98 <- cbind(tv98, "Drought" = rep("Non-drought", length(tv98$Butterfly)))
```

2001:

```{r 2001 formatting}

##2001

#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
tv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv01)[1] <- "1" #rename
tv01 <- tv01[, -1] #delete
#rename rows as sites
row.names(tv01) <- c("P01", "T01", "V01")
#removes Gnat column (not a comparable observation type)
tv01 <- tv01[ ,-c(2, 3)] 
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv01 <- cbind(tv01, "Butterfly" = c(0, 0, 0)) #add missing columns
tv01 <- tv01[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns
#add year
tv01 <- cbind(tv01, "Year" = rep("2001", length(tv01$Butterfly)))
#add drought
tv01 <- cbind(tv01, "Drought" = rep("Non-drought", length(tv01$Butterfly)))
```

2012:

```{r 2012 formatting}

##2012

#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
tv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv12)[1] <- "1" #rename
tv12 <- tv12[, -1] #delete
#rename rows as sites
row.names(tv12) <- c("APS", "APU", "BBB", "BL", "BB", "BV12", "bHL", "bHU", "ELL", "ELU", "EGH", "KP", "LLP1", "LLP2", "LLP3", "LLP4", "MFML", "MFMU", "ODR", "RPL", "RPU", "SVF", "SR", "UdH", "VF12", "VF212")
#removes Gnat column
tv12 <- tv12[,-c(3)]
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv12 <- cbind(tv12, "Nectar robber" = rep("0", length(tv12$Butterfly))) #add missing columns
tv12 <- tv12[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Small insect", "Solitary bee", "Wasp")] #order columns
#add year
tv12 <- cbind(tv12, "Year" = rep("2012", length(tv12$Butterfly)))
#add drought
tv12 <- cbind(tv12, "Drought" = rep("Drought", length(tv12$Butterfly)))
tv12 <- tv12[, -6] #delete Small Insect column

```

2018: 

```{r 2018 formatting}

##2018

#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
tv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv18)[1] <- "1" #rename
tv18 <- tv18[, -1] #delete
#rename rows as sites
row.names(tv18) <- c("AS", "BRG", "BBP", "BV", "ERO", "RP", "VF18")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv18 <- cbind(tv18, "Nectar robber" = rep("0", length(tv18$Butterfly)), "Solitary bee" = rep("0", length(tv18$Butterfly))) #add missing columns
tv18 <- tv18[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Small insect", "Solitary bee", "Wasp")] #order columns
#add year
tv18 <- cbind(tv18, "Year" = rep("2018", length(tv18$Butterfly)))
#add drought
tv18 <- cbind(tv18, "Drought" = rep("Drought", length(tv18$Butterfly)))
tv18 <- tv18[, -6] #delete
tv18 <- tv18[-2, ] #delete
```

The next step is to load all years together onto the same dataframe. This dataframe holds visitor data by site.

```{r}
#combine data
all <- rbind(tv97, tv98, tv01, tv12, tv18)
#remove descriptive columns
all.years <- all[,-c(8, 9)]
#make all column numeric
all.years[] <- lapply(all.years, as.numeric)
#transform into matrix
all.years <- as.matrix(all.years)
```

From my understanding of Adonis, a separate dataframe is required for the other data. At the moment, this includes Year and Drought information. 

```{r}
#create descriptive dataframe
all.des <- as.data.frame(all[,-c(1:7)])
#add site information
all.des <- cbind(all.des, c("P1", "T1", "V1", "P2", "T2", "V2", "P3", "T3", "V3", "APS", "APU", "Back of Baldy Bend", "Baldy Lunch", "Bellview Bench", "Bellview View", "billy's Hilly Lower", "billy's Hilly Upper", "Emerald Lake Lower", "Emerald Lake Upper", "End of Gothic Hill", "Kebler Pass", "Long Lake Parking Lot 1", "Long Lake Parking Lot 2", "Long Lake Parking Lot 3", "Long Lake Parking Lot 4", "Max Field Meadow Lower", "Max Field Meadow Upper", "Ol' Dusty Road", "Rosy Point Lower", "Rosy Point Upper", "Schofield/Vera Falls", "Slate River 2", "Up on dat hill", "Vera Falls", "Vera Falls 2", "Avery Slope", "Bellview Bench Park", "Bellview View2", "East River Overlook", "Rosy Point", "Vera Falls2"))
#name site column
colnames(all.des)[3] <- "Site"
```

Now that the data formatting is complete I can run the actual analysis. Here we are using Morisita-Horn.

```{r}
beta <- vegdist(all.years, method = "horn")
```


```{r}
adonis(beta ~ Drought, all.des, perm = 999)
```

