---
title: "All Ipo"
author: "Kelly Endres"
date: "April 17, 2020"
output: html_document
---
This R Markdown file contains all code for Endres et al. 20xx manuscript investigating the effects of drought on Ipomopsis aggregata niche breadth. We utilized Ipomopsis visitation data from several sites over five years, 1997, 1998, 2001, 2012, and 2018, with 2012 and 2018 determined to be drought years. Between years we compared visitor abundance, community change, and richness. We also analyzed weather data to confirm drought year determination. 

###Code Overview

1. Drought Analysis
+ Adonis
+ NMDS
2. Visitation Summary Statistics
3. Visitor Abundance
+ glmm abundance analysis
4. Richness
+ iNEXT analysis - Visit richness
+ glmm analysis - Visit richness
5. Community Change
+ Adonis analysis
+ NMDS



####Package Load and Data Import

```{r packages, include=FALSE}

#load packages 
library(glmmTMB)
library(fitdistrplus)
library(DHARMa)
library(ggplot2)
library(tidyverse)
library(vegan)
library(dplyr)
library(GGally)
library(reshape2)
library(iNEXT)
library(gridExtra)
library(viridis)
library(cowplot)
```

These analyses require the associated datasets:
gothicweather.csv
total visits per site per year by each visitor catagory.csv
visit.abundance.by.observation.day.csv

```{r}
weatherdf = read.csv("gothicweather.csv")

all= read.csv("total visits per site per year by each visitor catagory.csv")

all.years.ab= read.csv("visit.abundance.by.observation.day.csv")

```


### 1. Drought Analysis

Here we confirm that our catagorization of 2012 and 2018 as drought years is accurate 

```{r data import}

#import data
weatherdf<- read.csv("gothicweather.csv")

# water is in inches so convert to cm
weatherdf$Watercm= weatherdf$Water * 2.54

# remove winters before 1994/95
weatherdf = weatherdf[c(17:32,37:43),]

# if only 2012 and 2018 are classified as drought
weatherdf= weatherdf %>%
  mutate(Drought_New = as.factor(ifelse(Winter == "2011-12" | Winter== "2017-18", "Drought", "Non Drought")))


```

With the data loaded, it needs to be reformatted as a matrix. The non-numeric columns need to be removed, as they can't be used in the nmds. 

```{r matrix formatting}

#create matrix
mat <- as.matrix(weatherdf)

#rename rows
row.names(mat) <- weatherdf$Winter

#delete non-numeric columns
colnames(mat)[1] <- "1"
colnames(mat)[5] <- "5"
colnames(mat)[6] <- "6"
mat <- mat[, - c(1, 5, 6)]
```

The matrix needs to be formatted further to fit the form for the NMDS analysis. The drought column needs to be removed from the dataframe containing the rest of the data and put into its own value. 

```{r create df}

# Make dataframe
dmat <- as.data.frame(mat) 

#change columns to numeric
dmat$Total <- as.numeric(as.character(dmat$Total))
dmat$Apr.1 <- as.numeric(as.character(dmat$Apr.1))
dmat$May.1 <- as.numeric(as.character(dmat$May.1))
dmat$Water <- as.numeric(as.character(dmat$Water))
dmat$Above.freezing <- as.numeric(as.character(dmat$Above.freezing))
dmat$Days.covered <- as.numeric(as.character(dmat$Days.covered))
dmat$Drought <- as.numeric(as.character(dmat$Drought))

#create value with drought information
gr <- c(dmat$Drought)

#remove drought column from other data
colnames(dmat)[7] <- "7" #rename
dmat2 <- dmat[, -c(7:9)] #remove
```

Now that the data is formatted properly analysis can begin. We used analysis of variance with ADONIS and NMDS to visualize differences between drought and non-drought years. We expect that 2012 and 2018 will be drought years. 

# Adonis

Create a separate dataframe is of just Drought information. 


```{r}

des= weatherdf[,c(1, 12)] #remove everything but drought status

rownames(des) <- des[,1]
des[,1] <- NULL

```

Compute distances with Morisita-Horn

```{r}
dist.mat <- vegdist(dmat2, method = "horn")
```

Run analysis of variance with adonis2
```{r}

model= adonis2(dist.mat ~ Drought_New, des, perm = 999)

```

Drought is highly significant! Our drought categorization explains multivariate composites of many weather variables.

# NMDS
Visualize with the function metaMDS. 

```{r NMDS}

# run a NMDS ordination
ord <- metaMDS(dmat2)
```

Because we have a large dataset, the function automatically performs a square root transformation using Wisconsin standardization. 

Plot the NMDS 

```{r plotting}

plot(ord, disp="sites", type="n")
  ordihull(ord, gr, col=c("mediumturquoise", "indianred1"), lwd=3)
  ordiellipse(ord, gr, draw="polygon", col=c("mediumturquoise", "indianred1"), alpha = c(50, 50))
  orditorp(ord,display="species",col="black",air=0.01, cex = .8)
  # orditorp(ord,display="sites", select = sel, col= "black", air=0.01,cex=1)
  points(ord, display = "sites", cex = 0.5, pch = 3, col = "black")
plot(data.envfit, col = "black", cex = .5)
```

This visualization gives several important pieces of information. Green/blue specifies Non-drought and red is Drought. We see that the two areas show no overlap, indicating that the years are different in terms of weather. 

### 2. Visitation Summary Statistics


``` {r}
all= read.csv("total visits per site per year by each visitor catagory.csv")
```

```{r}
all.long= all %>% 
  pivot_longer(cols = Butterfly:Wasp ,names_to = "Species" , values_to = "total.visits")

Table1= merge(all.long %>% 
  group_by(Year, Species) %>% 
  summarise_at(vars(total.visits), funs(sum)) %>% 
  pivot_wider(names_from = Species, values_from = total.visits )
,
all.long %>% 
  group_by(Year) %>% 
  summarise_at(vars(total.visits), funs(sum))
, by = "Year")

Table1 %>%
    mutate_at(vars(c("Bumblebee", "Butterfly","Hoverfly","Hummingbird","Other.fly","Solitary.bee","Wasp")), 
               function (x) (x/Table1$total.visits)*100)

```

## 3. Visitor Abundance
+ glmm abundance analysis

###Abundance

With the following code we compare visitor abundance between years. This is important as we are not able to look at straight abundances between years due to numerous sampling effort differences. We want to compare these as it is an important aspect of how drought may impact plant-pollinator communities. 

To examine abundance differences and account for sampling effort differences, we will be using a linear mixed effect model, glmmTMB adding several different fixed and random effects. 

####Data formatting


```{r}
all.years.ab= read.csv("visit.abundance.by.observation.day.csv")
```

####Abundance Analysis

The first step to examine visitor abundance between drought and non-drought years is to determine the correct distribution for our data. It is most likely negative binomial, so we start by checking for that distribution.

```{r abundance distribution fit}

#fit data distribution
distcheck <- fitdist(all.years.ab$Visitor.Abundance, "nbinom")

#summarize nbinom fit
summary(distcheck)

#plot nbinom fit
plot(distcheck)
```

According the the plot of the distribution the data fits a negative binomial distribution. 

Now we are able to run our linear mixed effects model (glmmTMB) using a negative binomial distribution (nbinom2). Our response variable is given as total visitor abundance for each individual site visit. Drought, observer number, and amount of time observing are fixed effects, while site nested within year is a random effect. 

```{r abundance model}

#create model
abundancemod <- glmmTMB(Visitor.Abundance ~ Drought + Observers + Time + (1|Site/Year), all.years.ab, family = "nbinom2")

#summary
summary(abundancemod)
```

This indicates that the presence of drought impacts observed visitor abundance. The amount of time spent observing also impacts visitor abundance. 

```{r}
gg <- residuals(abundancemod)

plot(gg)

all.years.ab$residuals <- gg

all.years.ab$fitted <- abundancemod$fitted

simoutput <- simulateResiduals(fittedModel = abundancemod, n = 250)

plot(simoutput)
```


```{r}
all.years.ab$visithour <- all.years.ab$Visitor.Abundance/all.years.ab$Time
```


```{r}
# change factor levels for color plotting
all.years.ab$Drought <- factor(all.years.ab$Drought, levels = c("Non-drought", "Drought"))

visit=ggplot(all.years.ab, aes(x=Drought, y=Visitor.Abundance)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(colour = Drought), alpha = 0.5, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(y="Visitor Abundance") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)

```



## 4. Richness

# Summary stats richness

```{r}
# Just averages across year-site 
all.years.ab %>% 
  group_by(Year, Site) %>%
  tally(Richness)
  
all.years.ab %>% 
  group_by(Drought) %>%
  summarize(mean_rich= mean(Richness))

# Mean richness was1.750 in drought years and 0.697 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years.ab %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_rich= mean(Richness))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_rich= mean(site_rich))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_rich))

# changes values slightly
# Non-drought	0.6931687			
# Drought	1.4957265	

# Abundance the same way
# Just averages across year-site 
all.years %>% #CM CHANGE this to fit the name of the df in this markdown
  group_by(Year, Site) %>%
  tally(Visitor.Abundance)
  
all.years %>% 
  group_by(Drought) %>%
  summarize(mean_abund= mean(Visitor.Abundance))

# Mean abundance was 3.875 in drought years and 1.725 in non-drought years

# but this doesn't account for effort so mean of a mean of a mean might yield different results

# first average within site-year
mean= all.years.ab %>%
  group_by(Year, Site, Drought) %>%
  summarize(site_abud= mean(Visitor.Abundance))


year_mean= mean %>%
  group_by(Year, Drought) %>%
  summarize(year_abund= mean(site_abud))

year_mean %>%
  group_by(Drought) %>%
  summarize(grand_mean= mean(year_abund))

# changes values slightly
# Non-drought	1.670445			
# Drought	3.183226	
```

+ iNEXT analysis - Visit richness
+ glmm analysis - Visit richness

```{r}
test= as.data.frame(t(Table1))
names(test) <- test[1,]
test <- test[-1,]

#remove totals row
test=test[-8,]

# make values numeric
cols <- c(1:ncol(test)) 
test[,cols] <- as.numeric(as.character(unlist(stats[,cols])))

# change to a matrix
tv = as.matrix(test)
```


```{r drought dataframe}
tvdf <- as.data.frame(tv)

# FIGURE OUT WHY THESE VALUES ARE NOT THE SAME
two= tvdf %>%
  transmute(drought= tvdf$`2012` + tvdf$`2018`, nondrought= tvdf$`1997` + tvdf$`1998` +tvdf$`2001`)
```


# Discrepencies: 2012 solitary bees (33 vs 41). Totals for that year 736 in Table 1 in paper and 728 here

#####1997: 

```{r richness format 1997}

##1997

# format into total visits by site
#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
itv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(itv97) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(itv97)[1] <- "1" #rename
itv97 <- itv97[, -1] #delete
#transpose data
itv97 <- t(itv97) 

# create vector
#sum row totals in new dataframe
totals97 <- as.data.frame(rowSums(itv97))
#add 0s until totals97 has 9 rows total
totals97 <- rbind(totals97, list(0)) #wasp
totals97 <- rbind(totals97, list(0)) #other fly

#sort in decreasing order
totals97 <- sort(totals97$`rowSums(itv97)`, decreasing=TRUE)
```

#####1998: 

```{r richness format 1998}

##1998

# format into total visits by site
#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
itv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(itv98) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(itv98)[1] <- "1" #rename
itv98 <- itv98[, -1] #delete
#transpose data
itv98 <- t(itv98) 

# create vector
#sum row totals in new dataframe
totals98 <- as.data.frame(rowSums(itv98))
#add 0s until totals97 has 9 rows like the others 
totals98 <- rbind(totals98, list(0)) #wasp
totals98 <- rbind(totals98, list(0)) #butterfly

#sort in decreasing order
totals98 <- sort(totals98$`rowSums(itv98)`, decreasing=TRUE)

```

#####2001: 

```{r richness format 2001}

##2001

# format into toal visits by site
#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
itv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(itv01) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(itv01)[1] <- "1" #rename
itv01 <- itv01[, -1] #delete
#transpose data
itv01 <- t(itv01) 
itv01 <- itv01[-c(2),] #removes Gnat row
itv01 <- itv01[-c(2),] #removes Hawkmoth row

# create vector
#sum row totals in new dataframe
totals01 <- as.data.frame(rowSums(itv01))
#add 0s until totals01 has 9 rows like the others 
totals01 <- rbind(totals01, list(0)) #butterfly

#sort in decreasing order
totals01 <- sort(totals01$`rowSums(itv01)`, decreasing=TRUE)

```

#####2012:

```{r richness format 2012}

##2012

#format into total vists by site
#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#treat NAs as 1
ipo2012[is.na(ipo2012)] <- 1
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
itv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(itv12) <- c("Avery Picnic Slope", "Avery Picnic Upper", "Back of Baldy Bend", "Baldy Lunch", "Bellview Bench", "Bellview View", "billy's Hilly Lower", "billy's Hilly Upper", "Emerald Lake Lower", "Emerald Lake Upper", "End of Gothic Hill", "Kebler Pass", "Long Lake Parking Lot 1", "Long Lake Parking Lot 2", "Long Lake Parking Lot 3", "Long Lake Parking Lot 4", "Max Field Meadow Lower", "Max Field Meadow Upper", "Ol' Dusty Road", "Rosy Point Lower", "Rosy Point Upper", "Schofield/Vera Falls", "Slate River 2", "Up on dat hill", "Vera Falls", "Vera Falls 2")
#delete first column where sites were previously 
colnames(itv12)[1] <- "1" #rename
itv12 <- itv12[, -1] #delete

#transpose data
itv12 <- t(itv12) 
itv12 <- itv12[-c(3),] #removes Gnat column
itv12 <- itv12[-c(6),] #removes Small insect column

# create vector
#sum row totals in new dataframe
totals12 <- as.data.frame(rowSums(itv12))

#sort in decreasing order
totals12 <- sort(totals12$`rowSums(itv12)`, decreasing=TRUE)

```

#####2018:

```{r richness format 2018}

##2018

#format into total visits by site
#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
itv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(itv18) <- c("Avery Slope", "Behind Rustler's Gultch", "Bellview Bench Park", "Bellview View", "East River Overlook", "Rosy Point", "Vera Falls")
#delete first column where sites were previously 
colnames(itv18)[1] <- "1" #rename
itv18 <- itv18[, -1] #delete
#transpose data
itv18 <- t(itv18) 
itv18 <- itv18[-c(6),] #removes Small insect column

# create vector
#sum row totals in new dataframe
totals18 <- as.data.frame(rowSums(itv18))
#add 0s until totals18 has 9 rows like the others
totals18 <- rbind(totals18, list(0)) #solitary bee

#sort in decreasing order
totals18 <- sort(totals18$`rowSums(itv18)`, decreasing=TRUE)

```

Now that individual vectors are created for each year, we can combine them into one matrix. 

```{r total visitation}

#create total visitation matrix
itv <- cbind(totals97, totals98, totals01, totals12, totals18)
#rename columns by year
colnames(itv) <- c("1997", "1998", "2001", "2012", "2018")
```

Next the drought and non-drought data is combined into a different dataframe. From the "itv" dataframe that contains all groups two new dataframes are created that sum the drought data (2012 and 2018) and then the non-drought data (1997, 1998, 2001). 

```{r drought dataframe}

itvdf <- as.data.frame(itv)

##create drought dataframe
#sum column totals into new dataframe
drought <- transform(itvdf, drought = totals12 + totals18)
drought <- drought[, -c(1:5)]
drought <- as.data.frame(drought)

##create nondrought dataframe
#sum column totals into new dataframe
nondrought <- transform(itvdf, nondrought = totals97 + totals98 + totals01)
nondrought <- nondrought[, -c(1:5)]
nondrought <- as.data.frame(nondrought)

#combine into one dataframe
two <- cbind(drought, nondrought)
```


``` {r}

colnames(itv97) <- c("P97","T97", "V97")
Wasps= c(0,0,0)
Other_fly= c(0,0,0)
itv97=rbind(itv97,Wasps, Other_fly)
 rownames(itv97)[7] <- "Other fly"
 
 
colnames(itv98) <- c("P98","T98", "V98")
Butterfly = c(0,0,0)
itv98=rbind(itv98,Wasps, Butterfly)

colnames(itv01) <- c("P01","T01", "V901") # has diff numb of species
itv01= rbind(itv01, Butterfly)
rownames(itv01)[6] <- "Wasps"


colnames(itv18) <- paste(colnames(itv18), "18")

colnames(itv12) <- paste(colnames(itv12), "12")

# merge and the transpose into long dataframe

test=cbind(itv97, itv98, itv01)
test=merge(itv97,itv98, by="row.names",all.x=TRUE) # in differenet orders so need to deal with that first
rownames(test) <- test[,1]
test[,1] <- NULL

testtwo= merge(test, itv01, by="row.names", all.x=T)
row.names(testtwo)=testtwo[,1]
testtwo[,1] <- NULL
rownames(testtwo)[7] <- "Wasp"

testthree= merge(testtwo, itv18, by="row.names", all.x=T)
row.names(testthree)=testthree[,1]
testthree[,1] <- NULL

testfour= merge(testthree, itv12, by="row.names", all.x=T)
row.names(testfour)=testfour[,1]
testfour[,1] <- NULL


testfour[is.na(testfour)] <- 0
```
 
``` {r} 
##create drought dataframe
#sum column totals into new dataframe)
drought <- transform(tvdf, drought = totals12 + totals18)
drought <- drought[, -c(1:5)]
drought <- as.data.frame(drought)

##create nondrought dataframe
#sum column totals into new dataframe
nondrought <- transform(tvdf, nondrought = totals97 + totals98 + totals01)
nondrought <- nondrought[, -c(1:5)]
nondrought <- as.data.frame(nondrought)

#combine into one dataframe
two <- cbind(drought, nondrought)
```
### iNEXT analysis - Visit Abundance 

Next is the actual iNEXT analysis. We have decided to use only hill numbers 0 (species richness) 

#### by year

separate the years by diversity order (here just species richness). 


```{r analysis by diversity order}
#species richness
d1 <- iNEXT(tv, q=c(0), datatype="abundance", endpoint=NULL)

```


```{r plotting by diversity order}

# plot richness

g3 <- ggiNEXT(d1, type=1, grey=T) + 
  theme_classic() + 
  xlim(0,350) +
  ylim(0, 7.5) +
  scale_colour_manual(values=c("#00BFC4", "#00BFC4", "#00BFC4","#F8766D","#F8766D")) +
  scale_fill_manual(values=c("turquoise3", "turquoise3", "turquoise3", "indianred", "indianred")) +
  scale_shape_manual(values=c(15, 16, 17, 19, 18)) +
  labs(x = "Number of individual visits", title = "Richness", y = "Visitor category diversity") +
  labs(title = "(b)") +
  theme(axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title = element_text(hjust = 0.02, size= 18, face="bold"), legend.title=element_blank(),legend.text = element_text(size=8), legend.position = c(0.89,0.3)) + 
 guides(linetype = FALSE)

# change size of text
gb_3 <- ggplot_build(g3)
gb_3$data[[1]]$size <- 4
gt3 <- ggplot_gtable(gb_3)

```
# By Drought vs non-drought

```{r analyze by hill number}
i1 <- iNEXT(two, q=c(0), datatype="abundance", endpoint=NULL)
```

```{r}
# richness
order1 <- ggiNEXT(i1, type=1, grey=T) + 
  xlim(0,300) +
  ylim(0, 7.5) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.02, face = "bold", size= 18), axis.title = element_text(size=16), axis.text=element_text(size=14), legend.position = c(0.85,0.2), legend.title=element_blank()) + 
 guides(linetype = FALSE) +
  scale_colour_manual(values=c("#F8766D","#00BFC4")) +
  scale_fill_manual(values=c( "indianred1","turquoise3")) +
  labs(x = "Number of individual visits", y = "Visitor category diversity") +
   labs(title = "(a)")

gb_4 <- ggplot_build(order1)
gb_4$data[[1]]$size <- 4
gt4 <- ggplot_gtable(gb_4)


#graph together

panel= grid.arrange(gt4, gt3, ncol=2)
ggsave("Figure 2 Rarefied Richness.pdf", panel, device="pdf", width= 16.6, height = 12, units= "cm")
```


Richness analysis with glmm

```{r}
c <- fitdist(all.years.ab$Richness, "nbinom")

plot(c)

model3 <- glmmTMB(Richness ~ Drought + Observers + Time + (1|Site/Year), all.years.ab, family = "nbinom2")

summary(model3)

# check residuals
gg2 <- residuals(model3)
plot(gg2)


all.years.ab$residuals2 <- gg2

all.years.ab$fitted2 <- model3$fitted

simoutput2 <- simulateResiduals(fittedModel = model3, n = 250)
plot(simoutput2)

richness= ggplot(all.years.ab, aes(x=Drought, y=Richness)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_classic() + 
  labs(x="Drought", y="Richness") +
  theme(legend.position = "none", axis.title.x = element_blank(), axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title  = element_text(hjust = 0.02, face = "bold", size= 18)) +
  scale_color_viridis(discrete=TRUE)


```

Plot 1a and 1b in panel
```{r}
# use cowplot
panel=plot_grid(visit, richness, labels = c('A', 'B'))

ggsave("Figure 1 Abundance and Richness_revision", panel, device="pdf", width= 16.6, height = 14.0, units= "cm")
```


### 5. Community Change
+ Adonis analysis
+ NMDS


This set of analyses looks at differences in community composition between drought and non-drought years via dissimilarity using permutational multivariate analyses. This requires construction  of community similarity/dissimilarity matrices and conducting Adonis analyses, with drought vs. nondrought as the fixed effects. Adonis is useful for this analysis as it uses strata which takes into account random effects. We use the Morisita-Horn index with Adonis as it is insensitive to differences in sampling effort among categories. Therefore we can use raw data without correcting for sampling effort. 

####Data formatting

Load dataframe containing total visits per site by each visitor category. 

``` {r}
all= read.csv("total visits per site per year by each visitor catagory.csv")
```

Make df into a matrix
``` {r}
all= column_to_rownames(all, "Site")
#remove descriptive columns
all.years <- all[,-c(8, 9)]
#make all column numeric
all.years[] <- lapply(all.years, as.numeric)
#transform into matrix
all.years <- as.matrix(all.years)
```

To go with this dataframe we are creating a second dataframe that contains site, year, and drought information. 

```{r community year other variables dataframe}
#create descriptive dataframe
all.des <- as.data.frame(all[,-c(1:7)])

#create value with drought information
Drought <- c(all$Drought)
```

####Adonis analyses

Now that the data formatting is complete community composition analyses can be completed. Here we are using the Morisita-Horn method (horn) to examine visitation between years. 

```{r adonis analyses}

# adonis(beta ~ Drought, all.des, perm = 999)
adonis(all.years ~ Drought, all.des, perm = 100000, method= "horn")
```

Looking at the output of the permutational multivariate analysis, we do not find statistically significant differences in community composition (p > 0.05). 


####NMDS analysis
```{r NMDS}
# run NMDS ordination
nmds <- metaMDS(all.years, distance = "horn")
```


Analyzing code outputs:

```{r analyze output} 
#call function output
nmds
```

This gives us lots of information. We can see the transformations used on the data (wisconsin(sqrt)), the distance measure used (horn), number of dimensions (2), as well as other outputs. The output is automatically rotated so that axis 1 has larger explanatory power. Goodness of fit is given with stress. Here, stress = 0.18 indicating that the output is a fair match for this dataset. 

I also want to look at which variables are driving distance. 

```{r variables distance}
#pull nmds scores for visitor categories
visitor.cat.scores = as.data.frame(scores(nmds, "species"))  #Using the scores function from vegan to extract the species scores and convert to a data.frame
visitor.cat.scores$species = rownames(visitor.cat.scores)

#pull nmds scores for sites
data.scores = as.data.frame(scores(nmds))  #Using the scores function from vegan to extract the site scores and convert to a data.frame
data.scores$site = rownames(data.scores)  # create a column of site names, from the rownames of data.scores

data.scores.env= cbind(data.scores, all.des)

#reorder drought vs non-drought to flip the color scheme around
data.scores.env$Drought = relevel(factor(data.scores.env$Drought), ref = "Non-drought") 

```


Plot it
```{r plotting}

#plot NMDS
nmds.plot= ggplot() + 
  geom_point(aes(x=NMDS1, y=NMDS2, shape=Drought,color=Drought),size=3, data.scores.env) + # add the point markers
  geom_text(aes(x=NMDS1, y=NMDS2, label=species),alpha=0.5, visitor.cat.scores) +  # add the species labels
  # geom_text(aes(x=NMDS1, y=NMDS2, label=site),size=6,vjust=0, data.scores.env) +  # add the site labels
  scale_color_viridis(discrete=TRUE)+
  coord_equal() +
  theme_bw() + 
  theme(axis.text.x = element_blank(),  # remove x-axis text
        axis.text.y = element_blank(), # remove y-axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title.x = element_text(size=18), # remove x-axis labels
        axis.title.y = element_text(size=18), # remove y-axis labels
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank(),
        legend.position = "top", legend.title = element_blank(), legend.text = element_text(size= 12))

#achieves basically the same thing but with the ordiellipse
pdf('fig3.pdf')
  pl <- ordiplot(nmds, type = "none")
points(pl, "sites", pch=c(rep(17, 9), rep(19, 33)), col=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), bg=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), cex=1.2)
ordiellipse(nmds,groups=Drought,draw="polygon",col=c("#FDE725FF", "#440154FF"),label=F)
text(pl, "species", col="black", cex=1.2, font = 1)
dev.off()
```


 We see that the two areas show some overlap indicating partial similarities in composition confirmed by adonis. 




