---
title: "iNEXT analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

Here our goal is to analyse how drought impacts niche breadth changes though changes in species diversity. For this we will be using iNEXT. This package provides simple functions to compare and plot sample-size and coverage-based rarefaction and extrapolation of species diversity using individual based (abundance) data or sampling unit based (incident) data. Its important to use iNEXT as it uses rarefaction and takes into acount differences in sampling effort (of which there are many).However, due to differences in sampling effort between years that are difficult  to account for, we will be only preforming individual-based rarefaction for my Honors Thesis. Fortunatly, individual-based rarefaction tells what we want to know. 

### Code Overview

1. Package and Data Import
2. Data Formatting
- Individual Years
- Dataframe with all years
- Drought vs. Non-drought dataframe
3. iNEXT analysis
- all years
- Separating years by hill number
- 
Drought vs. Non-drought separated by hill number

iNEXT uses a particular data format for analysis. Visitation for each year is made into a vectar of total species abundance. The vector must be in order from low to high and have a number for each visitor catagory (9 total numbers). The first part of this code will be to make that format happen. 


```{r clear, include=FALSE}
#clear previous
rm(list = ls())
```

### Package and Data Import

```{r setup}

#load packages
library(dplyr)
library(tidyverse)
library(reshape2)
library(iNEXT)
library(gridExtra)
```

First we need to import the data. 

```{r data import}
#import data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```


### Data Formatting

#### Individual Years

Now that we have the data it needs to be formatted into the vectar format described above. An individual code chunk will be crated for each year. 

1997: 

```{r format 1997}

##1997

# format into total visits by site
#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
tv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv97) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv97)[1] <- "1" #rename
tv97 <- tv97[, -1] #delete
#transpose data
tv97 <- t(tv97) 

# create vector
#sum row totals in new dataframe
totals97 <- as.data.frame(rowSums(tv97))
#add 0s until totals97 has 9 rows total
totals97 <- rbind(totals97, list(0)) #wasp
totals97 <- rbind(totals97, list(0)) #nectar robber
totals97 <- rbind(totals97, list(0)) #other fly
totals97 <- rbind(totals97, list(0)) #small insect
#sort in decreasing order
totals97 <- sort(totals97$`rowSums(tv97)`, decreasing=TRUE)
```


1998: 

```{r format 1998}

##1998

# format into total visits by site
#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
tv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv98) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv98)[1] <- "1" #rename
tv98 <- tv98[, -1] #delete
#transpose data
tv98 <- t(tv98) 

# create vector
#sum row totals in new dataframe
totals98 <- as.data.frame(rowSums(tv98))
#add 0s until totals97 has 9 rows like the others 
totals98 <- rbind(totals98, list(0)) #wasp
totals98 <- rbind(totals98, list(0)) #nectar robber
totals98 <- rbind(totals98, list(0)) #butterfly
totals98 <- rbind(totals98, list(0)) #small insect
#sort in decreasing order
totals98 <- sort(totals98$`rowSums(tv98)`, decreasing=TRUE)

```

2001: 

```{r format 2001}

##2001

# format into toal visits by site
#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
tv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv01) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv01)[1] <- "1" #rename
tv01 <- tv01[, -1] #delete
#transpose data
tv01 <- t(tv01) 
tv01 <- tv01[-c(2),] #removes Gnat row
tv01 <- tv01[-c(2),] #removes Hawkmoth row

# create vector
#sum row totals in new dataframe
totals01 <- as.data.frame(rowSums(tv01))
#add 0s until totals01 has 9 rows like the others 
totals01 <- rbind(totals01, list(0)) #nectar robber
totals01 <- rbind(totals01, list(0)) #butterfly
totals01 <- rbind(totals01, list(0)) #small insect
#sort in decreasing order
totals01 <- sort(totals01$`rowSums(tv01)`, decreasing=TRUE)

```

2012:

```{r format 2012}

##2012

#format into total vists by site
#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#treat NAs as 1
ipo2012[is.na(ipo2012)] <- 1
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
tv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv12) <- c("Avery Picnic Slope", "Avery Picnic Upper", "Back of Baldy Bend", "Baldy Lunch", "Bellview Bench", "Bellview View", "billy's Hilly Lower", "billy's Hilly Upper", "Emerald Lake Lower", "Emerald Lake Upper", "End of Gothic Hill", "Kebler Pass", "Long Lake Parking Lot 1", "Long Lake Parking Lot 2", "Long Lake Parking Lot 3", "Long Lake Parking Lot 4", "Max Field Meadow Lower", "Max Field Meadow Upper", "Ol' Dusty Road", "Rosy Point Lower", "Rosy Point Upper", "Schofield/Vera Falls", "Slate River 2", "Up on dat hill", "Vera Falls", "Vera Falls 2")
#delete first column where sites were previously 
colnames(tv12)[1] <- "1" #rename
tv12 <- tv12[, -1] #delete
#transpose data
tv12 <- t(tv12) 
tv12 <- tv12[-c(3),] #removes Gnat column

# create vector
#sum row totals in new dataframe
totals12 <- as.data.frame(rowSums(tv12))
#add 0s until totals12 has 9 rows like the others
totals12 <- rbind(totals12, list(0)) #nectar robber
#sort in decreasing order
totals12 <- sort(totals12$`rowSums(tv12)`, decreasing=TRUE)

```

2018:

```{r format 2018}

##2018

#format into total visits by site
#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
tv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv18) <- c("Avery Slope", "Behind Rustler's Gultch", "Bellview Bench Park", "Bellview View", "East River Overlook", "Rosy Point", "Vera Falls")
#delete first column where sites were previously 
colnames(tv18)[1] <- "1" #rename
tv18 <- tv18[, -1] #delete
#transpose data
tv18 <- t(tv18) 

# create vector
#sum row totals in new dataframe
totals18 <- as.data.frame(rowSums(tv18))
#add 0s until totals18 has 9 rows like the others
totals18 <- rbind(totals18, list(0)) #solitary bee
totals18 <- rbind(totals18, list(0)) #Nectar robber
#sort in decreasing order
totals18 <- sort(totals18$`rowSums(tv18)`, decreasing=TRUE)

```

#### Dataframe with all years
Now that individual dataframes are created for each year, its necessary to combine them into one dataframe.

```{r total visitation}
#create total visitation dataframe
tv <- cbind(totals97, totals98, totals01, totals12, totals18)
#rename columns by year
colnames(tv) <- c("1997", "1998", "2001", "2012", "2018")

```

#### Drought vs. Non-drought dataframe

Next the drought and non-drought data must be combined into a different dataframe. From the "tv" dataframe that contains all groups two new dataframes will be created that sum the drought data (2012 and 2018) and then the non-drought data (1997, 1998, 2001). 

```{r drought dataframe}
tvdf <- as.data.frame(tv)

##create drought dataframe
#sum column totals into new dataframe
drought <- transform(tvdf, drought = totals12 + totals18)
drought <- drought[, -c(1:5)]
drought <- as.data.frame(drought)

##create nondrought dataframe
#sum column totals into new dataframe
nondrought <- transform(tvdf, nondrought = totals97 + totals98 + totals01)
nondrought <- nondrought[, -c(1:5)]
nondrought <- as.data.frame(nondrought)

#combine into one dataframe
two <- cbind(drought, nondrought)
```

### iNEXT analysis 

Next is the actual iNEXT analysis. We have decided to use only hill numbers 0 (species richness) and 1 (shannon's diversity). 

#### all years

The first step in iNEXT is to run the actual analysis, which then allows us to plot. 

```{r all iNEXT analysis}
#iNEXT analysis using individual based rarefaction
out <- iNEXT(tv, q=c(0, 1), datatype="abundance", endpoint=NULL)

```

This first plot gives us `q=0` and `q=1` for year each. Each is is on a separate graphs. 
```{r all graphing}
# R/E curves, separating by "site""
allplot <- ggiNEXT(out, type=1, facet.var="site") + 
  xlim(0,1000) +
  facet_wrap(~site, nrow = 3) +
  labs(title = "Niche Breadth Change",
       subtitle = "iNEXT",
       y = "Species Diversity", 
       x = "number of individuals") +
  theme_bw()
  
print(allplot) 
```

This graph gives us important information about each year. Notably, Extrapolation for 2012 doesn't begin until about 800 individuals. This graph will likely be included in the appendix section of the paper. 

*Further work is needed to make the graph look better. Namely the legend and axis ranges. 

#### Separating years by hill number

While the previous graph is important, for our paper it will be more useful to separate the years by diversity order (i.e. species richness and shannons). 


```{r analysis by diversity order}
#species richness
d1 <- iNEXT(tv, q=c(0), datatype="abundance", endpoint=NULL)

#shannons diversity 
d2 <- iNEXT(tv, q=c(1), datatype="abundance", endpoint=NULL)
```


```{r plotting by diversity order}
#plotting species richness 
g1 <- ggiNEXT(d1, type=1) + 
  xlim(0,350) +
  ylim(0, 8.5) +
  theme_bw() +
  scale_colour_manual(values=c("mediumturquoise", "mediumturquoise", "mediumturquoise", "indianred1", "indianred1")) +
  scale_fill_manual(values=c("turquoise3", "turquoise3", "turquoise3", "indianred1", "indianred1")) +
  scale_shape_manual(values=c(15, 16, 17, 19, 18)) +
  labs(x = "number of individuals", title = "richness") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) 
  
#plotting shannons diversity
g2 <- ggiNEXT(d2, type=1) + 
  xlim(0,350) +
  ylim(0, 8.5) +
  theme_bw() +
  scale_colour_manual(values=c("mediumturquoise", "mediumturquoise", "mediumturquoise", "indianred1", "indianred1")) +
  scale_fill_manual(values=c("turquoise3", "turquoise3", "turquoise3", "indianred1", "indianred1")) +
  scale_shape_manual(values=c(15, 16, 17, 19, 18)) +
  labs(x = "number of individuals", title = "Shannon") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) 
  
#plot both together
grid.arrange(g1, g2, ncol = 2)
```

Summary: These figures show higher diversity in the years 2018 and 2012. Utlizing a richness index (q=0), the three non-drought years (1997, 1998, 2001) show overlap, with slight overlap with 2018. Utlizing the Shannon's index (q=1), both 2012 and 2018 show higher diversity than the non-drought years.

Now that this is complete we will run an analysis with only two groups, drought and non-drought. This gets at our overall question in species richness difference between the two. Again, we expect higher species richness in drought years than non-drought years, indicating broader niche breadth. 

#### Drought vs. Non-drought

Now for the iNEXT analysis of drought vs. nondrought.

This first graph will show the rarefaction for each Drought and Non-drought separately 

```{r analyze drought and nondrought}
#iNEXT
out <- iNEXT(two, q=c(0, 1), datatype="abundance", endpoint=NULL)
```

```{r graphing drought and nondrought}
#graphs
final <- ggiNEXT(out, type=1, facet.var="site") + 
  xlim(0,300)
#plot
print(final)
```

Again, while this graph is important, it is more informative to look at the data separtated by hill numbers. 

#### Drought vs. Non-drought separated by hill number

```{r analyze by hill number}
i1 <- iNEXT(two, q=c(0), datatype="abundance", endpoint=NULL)
i2 <- iNEXT(two, q=c(1), datatype="abundance", endpoint=NULL)
```

```{r graphing by hill number}
# species richness
order1 <- ggiNEXT(i1, type=1) + 
  xlim(0,300) +
  ylim(0, 8) +
  theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "number of individuals") +
  labs(title = "richness")
 # scale_colour_manual(values=c("red", "blue")) 

gb4 <- ggplot_build(order1)
gb4$data[[2]]$size <- .5

#shannons diversity 
order2 <- ggiNEXT(i2, type=1) + 
  xlim(0,300) +
  ylim(0, 8) + 
  theme_bw() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  labs(x = "number of individuals") +
  labs(title = "Shannon")
 # scale_colour_manual(values=c("red", "blue")) 
  
gb3 <- ggplot_build(order1)
gb3$data[[2]]$size <- .5

#graph together
grid.arrange(order1, order2, ncol = 2)
```


Summary: This graph shows significant differences in diversity between drought and non-drought years using both richness (q=0) and Shannon's diversity (q=1). Greater species diversity was shown in drought years using both metrics.













