---
title: "Ipo reanalysis"
author: "Connor Morozumi"
date: "started 3.22.20; most recent edits `r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message=F)
```

```{r}
library(tidyverse)
library(glmmTMB)
library(fitdistrplus)
library(DHARMa)

library(iNEXT)
library(gridExtra)
library(reshape2)

```

Ipomopsis analysis questions

Compiling comments from Amy/Paul and Berry's thoughts via email. I see this as a list of possible reanalyses:

# TOC

1.	Run a glmm for visitor category richness as a response to better deal with lack of independence of site/years
2.	Include floral density as a covariate when analyzing visitor abundance
3.	How is rarefaction dealing with site pooling
4.	More direct linking of drought to reduced floral abundance at RMBL

# Dealing with non-independence

## Run a glmm for visitor category richness as a response to better deal with lack of independence of site/years

We had previously done this. I think it would be good to include. Amy suggests including both a glmm and rarefaction for richness. I don't see any issues really with that.

```{r}

#import data
ipo1997 <- read.csv("abundance_1997.csv") #1997
ipo1998 <- read.csv("abundance_1998.csv") #1998
ipo2001 <- read.csv("abundance_2001.csv") #2001
ipo2012 <- read.csv("abundance_2012.csv") #2012
ipo2018 <- read.csv("abundance_2018.csv") #2018

#'Year' in the 2012 dataset was actually entered as '2001'. Overwrote this
ipo2012$Year = 2012

```


```{r create dataframe}
# With the data imported we need to combine all the years into one dataframe for analysis. 
#combine data
all.years <- rbind(ipo1997, ipo1998, ipo2001, ipo2012, ipo2018)

#rename site column
colnames(all.years)[1] <- "Site"

all.years$Year =as.factor(all.years$Year)
```

```{r}
c <- fitdist(all.years$Richness, "nbinom")

plot(c)

model3 <- glmmTMB(Richness ~ Drought + Observers + Time + (1|Site/Year), all.years, family = "nbinom2")

summary(model3)

# check residuals
gg2 <- residuals(model3)
plot(gg2)


all.years$residuals2 <- gg2

all.years$fitted2 <- model3$fitted

simoutput2 <- simulateResiduals(fittedModel = model3, n = 250)
plot(simoutput2)

ggplot(all.years, aes(x=Drought, y=Richness)) +
  geom_boxplot(outlier.alpha  =  0, size = 0.5, width = 0.5) +
  geom_point(aes(color = Drought), alpha = 0.3, size = 3, position = position_jitter(w=0.2, h=0)) +
  theme_bw() + 
  labs(x="Drought", y="Richness")
```

Nesting Site and Year should deal with the issues of dependence.

Effect of drought is positively associated with higher richness (P= 0.002) after accounting for number of observers (ns) and amount of time observed (also significant, P= 0.023).

The sentence we had in old versions of the ms was:

Examining Visitor Richness yielded similar results, with both drought (p=0.00184) and time observing (p=0.0202) remaining significant. Like visitor abundance, visitor richness was higher in drought years compared to non-drought years.

## Site-based rarefactions
Berry says: We could also potentially do site-based rarefactions and see how they fall out graphically, if we always see that the drought ones for every site are above the non-drought ones for every site then that would be great.

I'm a bit confused, isn't this what we have in Fig 2A? The drought lines are higher, but 2018 has so much variability that we see overlap with the non-drought years. Because we sampled so much more in 2012 this effect is swamped out in the pooled approach (as it's currently visualized in the manuscript 2B).


```{r data import}
#import data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```


```{r format 1997}

##1997

# format into total visits by site
#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
tv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv97) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv97)[1] <- "1" #rename
tv97 <- tv97[, -1] #delete
#transpose data
tv97 <- t(tv97) 

# create vector
#sum row totals in new dataframe
totals97 <- as.data.frame(rowSums(tv97))
#add 0s until totals97 has 9 rows total
totals97 <- rbind(totals97, list(0)) #wasp
totals97 <- rbind(totals97, list(0)) #other fly

#sort in decreasing order
totals97 <- sort(totals97$`rowSums(tv97)`, decreasing=TRUE)
```



```{r format 1998}

##1998

# format into total visits by site
#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
tv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv98) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv98)[1] <- "1" #rename
tv98 <- tv98[, -1] #delete
#transpose data
tv98 <- t(tv98) 

# create vector
#sum row totals in new dataframe
totals98 <- as.data.frame(rowSums(tv98))
#add 0s until totals97 has 9 rows like the others 
totals98 <- rbind(totals98, list(0)) #wasp
totals98 <- rbind(totals98, list(0)) #butterfly

#sort in decreasing order
totals98 <- sort(totals98$`rowSums(tv98)`, decreasing=TRUE)

```



```{r format 2001}

##2001

# format into toal visits by site
#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
tv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv01) <- c("P", "T", "V")
#delete first column where sites were previously 
colnames(tv01)[1] <- "1" #rename
tv01 <- tv01[, -1] #delete
#transpose data
tv01 <- t(tv01) 
tv01 <- tv01[-c(2),] #removes Gnat row
tv01 <- tv01[-c(2),] #removes Hawkmoth row

# create vector
#sum row totals in new dataframe
totals01 <- as.data.frame(rowSums(tv01))
#add 0s until totals01 has 9 rows like the others 
totals01 <- rbind(totals01, list(0)) #butterfly

#sort in decreasing order
totals01 <- sort(totals01$`rowSums(tv01)`, decreasing=TRUE)

```



```{r format 2012}

##2012

#format into total vists by site
#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#treat NAs as 1
ipo2012[is.na(ipo2012)] <- 1
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
tv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv12) <- c("Avery Picnic Slope", "Avery Picnic Upper", "Back of Baldy Bend", "Baldy Lunch", "Bellview Bench", "Bellview View", "billy's Hilly Lower", "billy's Hilly Upper", "Emerald Lake Lower", "Emerald Lake Upper", "End of Gothic Hill", "Kebler Pass", "Long Lake Parking Lot 1", "Long Lake Parking Lot 2", "Long Lake Parking Lot 3", "Long Lake Parking Lot 4", "Max Field Meadow Lower", "Max Field Meadow Upper", "Ol' Dusty Road", "Rosy Point Lower", "Rosy Point Upper", "Schofield/Vera Falls", "Slate River 2", "Up on dat hill", "Vera Falls", "Vera Falls 2")
#delete first column where sites were previously 
colnames(tv12)[1] <- "1" #rename
tv12 <- tv12[, -1] #delete

#transpose data
tv12 <- t(tv12) 
tv12 <- tv12[-c(3),] #removes Gnat column
tv12 <- tv12[-c(7),] #removes Small insect column

# create vector
#sum row totals in new dataframe
totals12 <- as.data.frame(rowSums(tv12))

#sort in decreasing order
totals12 <- sort(totals12$`rowSums(tv12)`, decreasing=TRUE)

```


```{r format 2018}

##2018

#format into total visits by site
#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
tv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#rename rows as sites
row.names(tv18) <- c("Avery Slope", "Behind Rustler's Gultch", "Bellview Bench Park", "Bellview View", "East River Overlook", "Rosy Point", "Vera Falls")
#delete first column where sites were previously 
colnames(tv18)[1] <- "1" #rename
tv18 <- tv18[, -1] #delete
#transpose data
tv18 <- t(tv18) 
tv18 <- tv18[-c(6),] #removes Small insect column

# create vector
#sum row totals in new dataframe
totals18 <- as.data.frame(rowSums(tv18))
#add 0s until totals18 has 9 rows like the others
totals18 <- rbind(totals18, list(0)) #solitary bee

#sort in decreasing order
totals18 <- sort(totals18$`rowSums(tv18)`, decreasing=TRUE)

```


```{r total visitation}

#create total visitation dataframe
tv <- cbind(totals97, totals98, totals01, totals12, totals18)
#rename columns by year
colnames(tv) <- c("1997", "1998", "2001", "2012", "2018")
```

```{r drought dataframe}
tvdf <- as.data.frame(tv)

##create drought dataframe
#sum column totals into new dataframe
drought <- transform(tvdf, drought = totals12 + totals18)
drought <- drought[, -c(1:5)]
drought <- as.data.frame(drought)

##create nondrought dataframe
#sum column totals into new dataframe
nondrought <- transform(tvdf, nondrought = totals97 + totals98 + totals01)
nondrought <- nondrought[, -c(1:5)]
nondrought <- as.data.frame(nondrought)

#combine into one dataframe
two <- cbind(drought, nondrought)
```

```{r analysis by diversity order}
#species richness
d1 <- iNEXT(tv, q=c(0), datatype="abundance", endpoint=NULL)

```


```{r plotting by diversity order}

# plot richness

g3 <- ggiNEXT(d1, type=1, grey=T) + 
  theme_classic() + 
  xlim(0,350) +
  ylim(0, 7.5) +
  scale_colour_manual(values=c("#00BFC4", "#00BFC4", "#00BFC4","#F8766D","#F8766D")) +
  scale_fill_manual(values=c("turquoise3", "turquoise3", "turquoise3", "indianred", "indianred")) +
  scale_shape_manual(values=c(15, 16, 17, 19, 18)) +
  labs(x = "number of individual visits", title = "Richness", y = "visitor category diversity") +
  labs(title = "a)") +
  theme(axis.title = element_text(size=16), axis.text=element_text(size=14), plot.title = element_text(hjust = 0.02, size= 18, face="bold"), legend.title=element_blank(),legend.text = element_text(size=8), legend.position = c(0.89,0.3)) + 
 guides(linetype = FALSE)

# change size of text
gb_3 <- ggplot_build(g3)
gb_3$data[[1]]$size <- 4
gt3 <- ggplot_gtable(gb_3)

g3
```

# Include floral density as a covariate when analyzing visitor abundance

```{r}

# combine all years into one dataset

ipo1997_reduced= ipo1997 %>%
  dplyr::select(Year, Site, Flowers.per.Plant, plant., Visit.Date)  %>%
  rename(Plant.ID= plant.)
 
ipo2001_reduced= ipo2001 %>%
  dplyr::select(Year, Site, Flowers.per.Plant, Plant.ID, VisitDate) %>%
  rename(Visit.Date= VisitDate)

ipo2012_reduced= ipo2012 %>%
  dplyr::select(Year, Site, Flowers.per.Plant)
  
 ipo97and01=rbind(ipo1997_reduced, ipo2001_reduced)
  
 max= ipo97and01 %>%
  dplyr::select(Year, Site, Plant.ID, Flowers.per.Plant) %>%
  unite("Unique.ID", Site:Plant.ID, remove = FALSE) %>%
  arrange(Unique.ID) %>%
  arrange(desc(Flowers.per.Plant)) %>%
  group_by(Year, Site, Unique.ID) %>%
  slice(1)
 
 
```
To complete this analysis we would need to ask Mary for floral abundances. They had each plant individually tagged so it stands to reason that they do have these data. We have good density estimates for 2012 and 2018.

Here for the historic dataset we do have access to, I've summed Flowers per plant on all unique individuals where visitation was recorded, taking the maximum value for each unique individual since tagged plants often had repeat visits and their flower abundance varied through the season. For example, plant '21' at a given site had 2 flowers on July 2nd and 32 flowers on July 15th. But I don't think these are good proxy values for floral abundance since it is only the number of flowers on the plants that were visited. 

``` {r} 
 max %>%
  group_by(Year, Site) %>%
  tally(Flowers.per.Plant)

```

Abundances for 2012 and 2018

```{r results="hide"}

floral= read.csv("Ipomopsis 2012 floral data.csv")
floral18=read.csv("Ipomopsis 2018 floral data.csv")

#combine drought year datasets


floral= floral %>%
  filter(Flower.Species=="Ipomopsis aggregata")  %>%
  dplyr::select(Date, Site, Flower.Species, Total...of.Flowers, Individual.count) %>%
  rename(Total.flowers= Total...of.Flowers)

floral18= floral18 %>%
  filter(Species=="Ipomopsis aggregata")  %>%
  dplyr::select(Date, Site, Species, Flower.., Plant..) %>%
  rename(Total.flowers= Flower.., Individual.count=Plant.., Flower.Species= Species)

  
droughtyearfloral= rbind(floral, floral18)

# remove/ combine sites from 2012 that do not align with visitation data
droughtyearfloral$year <- as.factor(paste("20", sapply(as.character(droughtyearfloral$Date), function(x) unlist(strsplit(x, "/"))[3]), sep = ""))
levels(droughtyearfloral$year)

setdiff(floral %>% #see what sites don't overlap
  dplyr::select(Site) %>%
  group_by((Site)) %>%
  slice(1),

  floral18 %>%
  dplyr::select(Site) %>%
  group_by((Site)) %>%
  slice(1))


setdiff(floral %>% #see what sites don't overlap
  dplyr::select(Site) %>%
  group_by(Site) %>%
  slice(1), 
  
  df12%>%
  rename(Site=mdtsite) %>%
  dplyr::select(Site) %>%
  group_by(Site) %>%
  slice(1))
    
# Remove Judd, Slate River 1, remane Vera Falls 1 as just Vera Falls, rename belview to Bellview View
# not sure waht to do with Kebler and Schofield/Vera. Floral records have multiple sites e.g 1,2,3 but visitation only has 1

library(plyr)
droughtyearfloral$Site= revalue(droughtyearfloral$Site, c("Belview View"="Bellview View", "Vera Falls 1" = "Vera Falls"))

droughtyearfloral= droughtyearfloral %>%
  filter(Site != c("Judd Falls Maroon Sign", "Slate River 1"))
```

``` {r}
# find floral max for each site
droughtyearfloral  %>%
  dplyr::select(year, Site, Flower.Species, Total.flowers, Individual.count) %>%
  arrange(Site) %>%
  arrange(desc(Total.flowers)) %>%
  group_by(year, Site) %>%
  slice(1)
```


# How is rarefaction dealing with site pooling
Amy asks: Can we assume that data are pooled across sites for each year (fig. 2a) and then across all sites and years in 2b? 

Yes, this is the correct interpretation of Figure 2. 

Searching for ways rarefaction has been adjusted to deal with lack of independance... [Gotelli](https://www.uvm.edu/~ngotelli/manuscriptpdfs/Chapter%204.pdf) mentions that *sample based rarefaction* more effectively deals with the lack of independence of space and time between plots / transects/ etc. but I don't know if this quite does what we want.

# More direct linking of drought to reduced floral abundance at RMBL
Here I did a quick summary of our qnest data. Our qnest data suffers from unequal sampling through the years (eg. We sampled twice as much in the drought year 2018 compared to the other years). Additionally, I have some worries about accuracy through the years in terms of what we counted, especially 2016. But here are a few ways to pull out a signal.

```{r load dataset}

flowers= read.csv("qnest_plantsurvey.csv", header = T)

#check for blank values

# summary(flowers)


flowers= flowers %>%
  filter(plant_name != "None"| plant_name != "")

flowers= flowers %>%
  filter(total_flowers != "NA")  %>%
   filter(date != "")


#change to factors if applicable
flowers$segment= as.factor(flowers$segment)

# check for spelling errors within plant species
# levels(flowers$plant_name)

flowers= flowers %>%
  filter(plant_name != "")

levels(flowers)[levels(flowers)=="Fringed mustard"] <- "Descurainia californica"
levels(flowers)[levels(flowers)=="vicia americana"] <- "Vicia americana"

library(plyr)
flowers$plant_name= revalue(flowers$plant_name, c("Fringed mustard "="Descurainia californica", "vicia americana"= "Vicia americana", "Potentilla fruiticosa"= "Potentilla fruticosa", "Arctostaphylos uva-ursi" ="Vaccinium caespitosum", "helianthella quinquenervis" = "Helianthella quinquenervis", "LInum lewisii" = "Linum lewisii", "Taraxacum officianale"= "Taraxacum officinale", "Aquilegia coerulea"= "Aquilegia caerulea"))


```

```{r}
# make a year column
#compute year variable and append to the end of site names
flowers$year <- as.factor(paste("20", sapply(as.character(flowers$date), function(x) unlist(strsplit(x, "/"))[3]), sep = ""))

# levels(flowers$year)


#add permenant plot identifiers
# if combo == starts with the string xx and contains string manip & year == 2017
# if combo == starts with the string xx and contains string 

```


Summarize by site and year

```{r}

samples= flowers %>%
  dplyr::select (c(site, date, total_flowers, year)) %>%
  group_by(site, date, year) %>%
  tally(total_flowers)

as.tibble(samples)
```

5 samples each in 2016, 5 AP and VB, 7 GT 2017, 9 samples GT and AP and 6 at VB in 2018, 6 at all sites 2019 

No correction for number of samples
```{r summarize}

count= flowers %>%
  dplyr::select (c(site, total_flowers, year)) %>%
  group_by(site, year) %>%
  tally(total_flowers)

as.tibble(count)

by_species= flowers %>%
  dplyr::select (c(site, plant_name, total_flowers, year, date)) %>%
  group_by(site, year, plant_name, date) %>%
  tally(total_flowers)
# 
# flowers %>%
#   filter(grepl('Ach', plant_name)) %>%
#   dplyr::select (c(site, total_flowers, year)) %>%
#   group_by(site, year) %>%
#   tally(total_flowers)

```

Correcting for number of samples

To deal with the fact that we sampled more in certain years, I just took the cumulative floral peak each year. 

1. First just by filtering for the week with the highest count

*Peak abundances – roughly a 50% reduction in floral abundances when compared to averaged 2017, 2019 peaks at the same site *
```{r}
peaks= samples %>%
  group_by(site, year) %>%
  filter(n == max(n))

ggplot(peaks, aes(year, n, colour = site))+
  geom_jitter( height=NULL, width = 0.25, size= 3) +
  theme_classic()

```

2. By filtering just for the peak total abundance across all species we miss lots of info. Here I've filtered by peak abundances for each species at a time then summed

```{r}
by_species= by_species %>%
  group_by(site, year, plant_name) %>%
  filter(n == max(n))
  
  
spec_peak_counts= by_species  %>%
  group_by(site, year) %>%
  tally(n)

ggplot(spec_peak_counts, aes(year, n, colour = site))+
  geom_jitter( height=NULL, width = 0.25, size= 3) +
  theme_classic()

```



