---
title: "Ipomopsis cleaning and data prep doc"
output: html_notebook
---

 This Rmarkdown file is associated with Endres et al. 20xx manuscript on Ipomopsis niche breadth responses to drought perterbation. This creates the following datasets that are used in the Rmarkdown file (Endres et al. Ipomopsis data analysis) to be publically available:
 1. visitation by visitor category in all years. Used in glmm abundance analysis.
 2. 

1. Visitation by visitor category in all years

First step is to import the data needed for this code. I manually reformated the data in excel as it would take much less time and was more straightforward. These new data sheets have been uploaded separately into GitHub.

```{r data inport}

#inport data
ipo1997 <- read.csv("abundance_1997.csv") #1997
ipo1998 <- read.csv("abundance_1998.csv") #1998
ipo2001 <- read.csv("abundance_2001.csv") #2001
ipo2012 <- read.csv("abundance_2012.csv") #2012
ipo2018 <- read.csv("abundance_2018.csv") #2018
```

With the data imported we need to combine all the years into one dataframe for analysis. 

```{r create dataframe}
#combine data
all.years <- rbind(ipo1997, ipo1998, ipo2001, ipo2012, ipo2018)

#rename site column
colnames(all.years)[1] <- "Site"

```

Export data
```{r}
write.csv(file= "Ipomopsis visitation by visitor category all years.csv", all.years)
```



Investigating model output glmmTMB

Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting') 
Following [this vignette](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html)

```{r}
fixef(abundancemod)
```

