---
title: "ipoweather"
author: "Kelly Endres"
date: "November 20, 2019"
output: html_document
---

In this code the goal is to determine the significance of drought over different years. We will then be able to determine which years to include as drought and which to include as non-drought in our analysis. The last column, drought, was created to group years into drought and non-drought for the NMDS analysis. I defined drought as anything with a snowfall total below 700cm, which was very low compared to average snowfall. 

```{r packages, include=FALSE}

#load packages
library(vegan)
library(dplyr)
library(GGally)
library(tidyverse)
```

Data is imported from an excel file containing various weather data points from Gothic, obtained from billy barr's website. 

```{r data inport}

#inport data
weatherdf<- read.csv("gothicweather.csv")

# water is in inches so convert to cm
weatherdf$Watercm= weatherdf$Water * 2.54


# if only 2012 and 2018 are classified as drought
weatherdf= weatherdf %>%
  mutate(Drought_New = ifelse(Winter == "2011-12" | Winter== "2017-18", "Drought", "Non Drought"))


```

Initial determination based on difference from long term means
```{r}
mean(weatherdf$Total) # 1066.651
round(mean(weatherdf$Total) -(1.5*sd(weatherdf$Total)), 0)

650/ mean(weatherdf$Total) 

1.5*sd(weatherdf$Total) #568.41
2*sd(weatherdf$Watercm) #37.66738

weatherdf %>% 
  mutate(is.sd= ifelse(Total < 700, "less", "no"))

weatherdf %>% 
  mutate(is.sd= ifelse(Total <= round(mean(weatherdf$Total) -(1.5*sd(weatherdf$Total)), 0), "less", "no"))
```


For subsequent analyses let's only use data from the study period

```{r}
# remove winters before 1994/95
weatherdf = weatherdf[c(17:32,37:43),]
```


With the data loaded, it needs to be reformatted as a matrix. The non-numeric columns need to be removed, as they can't be used in the nmds. 

```{r matrix formatting}

#create matrix
mat <- as.matrix(weatherdf)

#rename rows
row.names(mat) <- weatherdf$Winter

#delete non-numeric columns
colnames(mat)[1] <- "1"
colnames(mat)[5] <- "5"
colnames(mat)[6] <- "6"
mat <- mat[, - c(1, 5, 6)]
```

The matrix needs to be formatted further to fit the form for the NMDS analysis. The drought column needs to be removed from the dataframe containing the rest of the data and put into its own value. 

```{r create df}

# Make dataframe
dmat <- as.data.frame(mat) 

#change columns to numeric
dmat$Total <- as.numeric(as.character(dmat$Total))
dmat$Apr.1 <- as.numeric(as.character(dmat$Apr.1))
dmat$May.1 <- as.numeric(as.character(dmat$May.1))
dmat$Water <- as.numeric(as.character(dmat$Water))
dmat$Above.freezing <- as.numeric(as.character(dmat$Above.freezing))
dmat$Days.covered <- as.numeric(as.character(dmat$Days.covered))
dmat$Drought <- as.numeric(as.character(dmat$Drought))

#create value with drought information
gr <- c(dmat$Drought)

#remove drought column from other data
colnames(dmat)[7] <- "7" #rename
dmat2 <- dmat[, -c(7:9)] #remove
```

Now that the data is formatted properly analysis can begin. Again, I will be using NMDS to visualize differences between drough and non-drought years. We expect that 2012 and 2018 will be drought years. 

Here I will use function metaMDS to preform the analysis. 

```{r NMDS}

# run a NMDS ordination
ord <- metaMDS(dmat2)
```

Because we have a large dataset, the function automatically preforms a square root transformation using Wisconsin standardation. 

We will be analysing the NMDS via plotting. 

```{r plotting}

plot(ord, disp="sites", type="n")
  ordihull(ord, gr, col=c("mediumturquoise", "indianred1"), lwd=3)
  ordiellipse(ord, gr, draw="polygon", col=c("mediumturquoise", "indianred1"), alpha = c(50, 50))
  orditorp(ord,display="species",col="black",air=0.01, cex = .8)
  # orditorp(ord,display="sites", select = sel, col= "black", air=0.01,cex=1)
  points(ord, display = "sites", cex = 0.5, pch = 3, col = "black")
plot(data.envfit, col = "black", cex = .5)
```

This visualization gives several important pieces of information. Green/blue specifies Non-drought and red is Drought. We see that the two areas show no overlap, indicating that the years are different in terms of weather. 

We can also fit our environmental variable, drought, onto the ordination. This visualizes how the differences in weather between years are impacted by drought. It will also give us what percent of the variation can be explained by drough categorization.

```{r}
data.envfit <- envfit(ord, gr, lwd = 2)
plot(data.envfit, col = "black", cex = .5)
data.envfit
```

# Adonis

From my understanding of Adonis, a separate dataframe is required for the other data. At the moment, this includes just Drought information. 


```{r}
des= weatherdf[,c(1, 10)] #remove everything but drought status

des$Drought= as.factor(ifelse(des$Drought == 1, "Non Drought", "Drought"))
rownames(des) <- des[,1]
des[,1] <- NULL


# if only classifying 2012 and 2018 as drought

des= weatherdf[,c(1, 12)] #remove everything but drought status

rownames(des) <- des[,1]
des[,1] <- NULL


```

Now that the data formatting is complete I can run the actual analysis. Here we are using Morisita-Horn.

```{r}
dist.mat <- vegdist(dmat2, method = "horn")
```


```{r}
adonis(dist.mat ~ Drought, des, perm = 999)

#only 2012 and 2018 classified as drought
model= adonis(dist.mat ~ Drought_New, des, perm = 999)

```

Drought is highly significant! Our arbitrary drought catagorization explains multivariate composites of many weather variables.
