---
title: "Drought NMDS"
author: "Kelly Endres"
date: "January 23, 2019"
output:
  html_document: default
  pdf_document: default
---

This code impliments non-metric multi-dimensional scaling (NMDS) to visualize community composition changes between drought and non-drought years. NMDS is an ordination technique used for non-linear data (e.g. species abundance counts) that represents the position of data in multidimensional space using distances between points for ordination.

The first two code chunks provide background set up and load needed packages. 

```{r setup, include=FALSE}
#clear environment
rm(list = ls())
```

```{r packages, include=FALSE}

#load packages
library(vegan)
library(dplyr)
library(tidyverse)
library(reshape2)
library(GGally)
```

Data is imported from excel files which contain visitation data by visitor category, as well as site, year, and drought information. 

```{r data inport}

#inport data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```

Now that the data is loaded, each year needs to be reformated for the NMDS. Because of some differences between how the data sheets are set up (due both to differences in related information collected and my initial lack of experience with data cleaning) the exact formating for each year varies. Ultimately, each year will have the same format so one master dataframe can be made. 

One code chunk is used to format each year. 

1997: 

```{r 1997 formating}

##1997

#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits (removing extra information)
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
tv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv97)[1] <- "1" #rename
tv97 <- tv97[, -1] #delete
#rename rows as sites
row.names(tv97) <- c("P97", "T97", "V97")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv97 <- cbind(tv97, "Other fly" = c(0, 0, 0), "Wasp" = c(0, 0, 0)) #add missing columns
tv97 <- tv97[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns alphabetically
#add year column
tv97 <- cbind(tv97, "Year" = rep("1997", length(tv97$Butterfly)))
#add drought column
tv97 <- cbind(tv97, "Drought" = rep("Non-drought", length(tv97$Butterfly)))
```

1998: 

```{r 1998 formatting}

##1998

#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
tv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv98)[1] <- "1" #rename
tv98 <- tv98[, -1] #delete
#rename rows as sites
row.names(tv98) <- c("P98", "T98", "V98")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv98 <- cbind(tv98, "Butterfly" = c(0, 0, 0), "Wasp" = c(0, 0, 0)) #add missing columns
tv98 <- tv98[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns alphabetically
#add year
tv98 <- cbind(tv98, "Year" = rep("1998", length(tv98$Butterfly)))
#add drought
tv98 <- cbind(tv98, "Drought" = rep("Non-drought", length(tv98$Butterfly)))
```

2001:

```{r 2001 formatting}

##2001

#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
tv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv01)[1] <- "1" #rename
tv01 <- tv01[, -1] #delete
#rename rows as sites
row.names(tv01) <- c("P01", "T01", "V01")
#removes Gnat column (not a comparable observation type)
tv01 <- tv01[ ,-c(2, 3)] 
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv01 <- cbind(tv01, "Butterfly" = c(0, 0, 0)) #add missing columns
tv01 <- tv01[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Solitary bee", "Wasp")] #order columns
#add year
tv01 <- cbind(tv01, "Year" = rep("2001", length(tv01$Butterfly)))
#add drought
tv01 <- cbind(tv01, "Drought" = rep("Non-drought", length(tv01$Butterfly)))
```

2012:

```{r 2012 formatting}

##2012

#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
tv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv12)[1] <- "1" #rename
tv12 <- tv12[, -1] #delete
#rename rows as sites
row.names(tv12) <- c("APS", "APU", "BBB", "BL", "BB", "BV12", "bHL", "bHU", "ELL", "ELU", "EGH", "KP", "LLP1", "LLP2", "LLP3", "LLP4", "MFML", "MFMU", "ODR", "RPL", "RPU", "SVF", "SR", "UdH", "VF12", "VF212")
#removes Gnat column
tv12 <- tv12[,-c(3)]
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv12 <- cbind(tv12, "Nectar robber" = rep("0", length(tv12$Butterfly))) #add missing columns
tv12 <- tv12[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Small insect", "Solitary bee", "Wasp")] #order columns
#add year
tv12 <- cbind(tv12, "Year" = rep("2012", length(tv12$Butterfly)))
#add drought
tv12 <- cbind(tv12, "Drought" = rep("Drought", length(tv12$Butterfly)))
tv12 <- tv12[, -6] #delete Small Insect column

```

2018: 

```{r 2018 formatting}

##2018

#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
tv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv18)[1] <- "1" #rename
tv18 <- tv18[, -1] #delete
#rename rows as sites
row.names(tv18) <- c("AS", "BRG", "BBP", "BV", "ERO", "RP", "VF18")
#Not every type of visitor was observed, and these do not have columns. Columns for missing visitors need to be added so that the years can be combined. 
tv18 <- cbind(tv18, "Nectar robber" = rep("0", length(tv18$Butterfly)), "Solitary bee" = rep("0", length(tv18$Butterfly))) #add missing columns
tv18 <- tv18[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Other fly", "Small insect", "Solitary bee", "Wasp")] #order columns
#add year
tv18 <- cbind(tv18, "Year" = rep("2018", length(tv18$Butterfly)))
#add drought
tv18 <- cbind(tv18, "Drought" = rep("Drought", length(tv18$Butterfly)))
tv18 <- tv18[, -6] #delete
tv18 <- tv18[-2, ] #delete
```


The next step is to load all years together onto the same dataframe. The data is organized by site, then by year. 

```{r create total df}

#combine data
all.years <- rbind(tv97, tv98, tv01, tv12, tv18)
#remove drought column making new dataframe
all.tv <- all.years[, -c(8, 9)]
#make dataframe numeric
all.tv[] <- lapply(all.tv, as.numeric)
#create value with drought information
Drought <- c(all.years$Drought)
```

Now that the data is formatted properly analysis can begin. Again, I will be using NMDS to visualize differences in community composition. We expect that drought will be a significant driver of changes in community composition. 

Here I will use function metaMDS. This function is commonly used for ecological count data. Similar to my adonis analysis, I will run the function using the Morisita-Horn index due to its insensitivity to sampling effort differences. 

```{r NMDS}
# run NMDS ordination
nmds <- metaMDS(all.tv, distance = "horn")
```

Because we have a large dataset, the function automatically preforms a square root transformation using Wisconsin standardation. 

NMDS can be analyzed using the code outputs and by plotting. I will be doing both here. 

Analyzing code outputs:

```{r analyze output} 
#call function output
nmds
```

This gives us lots of information. We can see the transformations used on the data (wisconsin(sqrt)), the disntance measure used (horn), number of dimensions (2), as well as other outputs. The output is automatically rotated so that axis 1 has larger explanitory power. Goodness of fit is given with stress. Here, stress = 0.18 indicating that the output is a fair match for this dataset. 

I also want to look at which variables are driving distance. 

```{r variables distance}
#examine species
nmds$species
#examine points
nmds$points
```

These specific outputs can help us examine the ordination when plotted. For plotting variables are grouped by drought and non-drought, visualized with the colored polygons. The polygons represent (WHAT EXACTLY??). 

```{r plotting}

#plot NMDS
ordiplot(nmds,type="n") + # no points 
  theme_bw()
  orditorp(nmds,display="species",col="black",air=0.01, cex = .6) #add speices
  orditorp(nmds,display="sites", col=c(rep("turquoise4", 9), rep("firebrick", 33)), cex=.8, air=0.01) #add sites
  ordiellipse(nmds, Drought, draw = "polygon", col=c("mediumturquoise", "indianred1"), kind = "sd", alpha = c(50, 50)) #add overlap circles 

ordiplot(nmds,type="n") + # no points 
  theme_bw()
  orditorp(nmds,display="species",col="black",air=1, cex = 1.5) #add speices
  orditorp(nmds,display="sites", col=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), cex=1, air=200, pch= 19) #add sites
  ordiellipse(nmds, Drought, draw = "polygon", col=c("#440154FF", "#FDE725FF"), kind = "sd", alpha = c(50, 50)) #add overlap circles   

  pdf('fig3.pdf')
  pl <- ordiplot(nmds, type = "none")
points(pl, "sites", pch=c(rep(17, 9), rep(19, 33)), col=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), bg=c(rep("#440154FF", 9), rep("#FDE725FF", 33)), cex=1.2)
ordiellipse(nmds,groups=Drought,draw="polygon",col=c("#FDE725FF", "#440154FF"),label=F)
text(pl, "species", col="black", cex=1.2, font = 1)
dev.off()

```


This visualization g ives several important pieces of information. Green/blue specifies non-drought and red is Drought for both the circles and the sites. We see that the two areas show some overlap intidating partial similarities in composition. 

We can also fit our environmental variable, drought, onto the ordination. This visualizes how drought is impacting the spread of sites and species. It will also give us what percent of the variation can be explained by drough categorization. 

```{r environment}
data.envfit <- envfit(nmds,Drought, lwd = 2)
plot(data.envfit, col = "black", cex = .5)
data.envfit
```

