---
title: "Drought NMDS"
author: "Kelly Endres"
date: "January 23, 2019"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
```

```{r packages, include=FALSE}
#load packages
library(vegan)
library(dplyr)
library(tidyverse)
library(reshape2)
library(GGally)
```

```{r data inport, include=FALSE}
#inport data
ipo1997 <- read.csv("data_1997.csv") #1997
ipo1998 <- read.csv("data_1998.csv") #1998
ipo2001 <- read.csv("data_2001.csv") #2001
ipo2012 <- read.csv("data_2012.csv") #2012
ipo2018 <- read.csv("data_2018.csv") #2018
```

To complete NMDS, the data from all years must be together on the same matrix. This requires loading the correct data and formatting it the same. In particular, the visitor column must be in the same order. 


```{r 1997 formating}
##1997
#treat year as factor
ipo1997$Year <- as.factor(ipo1997$Year)
#create dataframe with site, visitor, and total visits
df97 <- cbind.data.frame(mdtsite = ipo1997$Site, mdtvisitor = ipo1997$Visitor, total.vis = ipo1997$Total.Visits)
#organize data by site then visitor
tv97 <- dcast(df97, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv97)[1] <- "1" #rename
tv97 <- tv97[, -1] #delete
#rename columns
row.names(tv97) <- c("P97", "T97", "V97")
#add missing columns
tv97 <- cbind(tv97, "Other fly" = c(0, 0, 0), "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Wasp" = c(0, 0, 0))
#order columns
tv97 <- tv97[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
tv97 <- cbind(tv97, "Year" = c(1997, 1997, 1997))
#add drought
tv97 <- cbind(tv97, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))
```


```{r 1998 formatting}
##1998
#treat year as factor
ipo1998$Year <- as.factor(ipo1998$Year)
#create dataframe with site, visitor, and total visits
df98 <- cbind.data.frame(mdtsite = ipo1998$Site, mdtvisitor = ipo1998$Visitor, total.vis = ipo1998$Total.Visits)
#organize data by site then visitor
tv98 <- dcast(df98, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv98)[1] <- "1" #rename
tv98 <- tv98[, -1] #delete
#rename columns
row.names(tv98) <- c("P98", "T98", "V98")
#add missing columns
tv98 <- cbind(tv98, "Butterfly" = c(0, 0, 0), "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Wasp" = c(0, 0, 0))
#order columns
tv98 <- tv98[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
tv98 <- cbind(tv98, "Year" = c(1998, 1998, 1998))
#add drought
tv98 <- cbind(tv98, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))
```


```{r 2001 formatting}
##2001
#treat year as factor
ipo2001$Year <- as.factor(ipo2001$Year)
#create dataframe with site, visitor, and total visits
df01 <- cbind.data.frame(mdtsite = ipo2001$Site, mdtvisitor = ipo2001$Visitor, total.vis = ipo2001$Total.Visits)
#organize data by site then visitor
tv01 <- dcast(df01, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv01)[1] <- "1" #rename
tv01 <- tv01[, -1] #delete
#rename columns
row.names(tv01) <- c("P01", "T01", "V01")
#removes Gnat  andda
tv01 <- tv01[ ,-c(2, 3)] 
#add missing columns
tv01 <- cbind(tv01, "Nectar robber" = c(0, 0, 0), "Small insect" = c(0, 0, 0), "Butterfly" = c(0, 0, 0))
#order columns
tv01 <- tv01[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
tv01 <- cbind(tv01, "Year" = c(2001, 2001, 2001))
#add drought
tv01 <- cbind(tv01, "Drought" = c("Non-drought", "Non-drought", "Non-drought"))
```


```{r 2012 formatting}
##2012
#treat year as factor
ipo2012$Year <- as.factor(ipo2012$Year)
#treat NAs as 1
ipo2012[is.na(ipo2012)] <- 1
#create dataframe with site, visitor, and total visits
df12 <- cbind.data.frame(mdtsite = ipo2012$Site, mdtvisitor = ipo2012$Visitor, total.vis = ipo2012$Total.Visits)
#organize data by site then visitor
tv12 <- dcast(df12, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv12)[1] <- "1" #rename
tv12 <- tv12[, -1] #delete
#rename column
row.names(tv12) <- c("APS", "APU", "BBB", "BL", "BB", "BV12", "bHL", "bHU", "ELL", "ELU", "EGH", "KP", "LLP1", "LLP2", "LLP3", "LLP4", "MFML", "MFMU", "ODR", "RPL", "RPU", "SVF", "SR", "UdH", "VF12", "VF212")
#removes Gnat column
tv12 <- tv12[,-c(3)]
#add missing columns
tv12 <- cbind(tv12, "Nectar robber" = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0))
#order columns
tv12 <- tv12[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
tv12 <- cbind(tv12, "Year" = c(2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012, 2012))
#add drought
tv12 <- cbind(tv12, "Drought" = c("Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought"))
```


```{r 2018 formatting}
##2018
#treat year as factor
ipo2018$Year <- as.factor(ipo2018$Year)
#create dataframe with site, visitor, and total visits
df18 <- cbind.data.frame(mdtsite = ipo2018$Site, mdtvisitor = ipo2018$Visitor, total.vis = ipo2018$Total.Visits)
#organize data by site then visitor
tv18 <- dcast(df18, mdtsite ~ mdtvisitor, sum)
#delete site column
colnames(tv18)[1] <- "1" #rename
tv18 <- tv18[, -1] #delete
#rename columns
row.names(tv18) <- c("AS", "BRG", "BBP", "BV", "ERO", "RP", "VF18")
#add missing columns
tv18 <- cbind(tv18, "Nectar robber" = c(0, 0, 0, 0, 0, 0, 0), "Solitary bee" = c(0, 0, 0, 0, 0, 0, 0))
#order columns
tv18 <- tv18[ , c("Butterfly", "Bumblebee", "Hoverfly", "Hummingbird", "Nectar robber", "Other fly", "Small insect", "Solitary bee", "Wasp")]
#add year
tv18 <- cbind(tv18, "Year" = c(2018, 2018, 2018, 2018, 2018, 2018, 2018))
#add drought
tv18 <- cbind(tv18, "Drought" = c("Drought", "Drought", "Drought", "Drought", "Drought", "Drought", "Drought"))
```


The next step is to load all years together onto the same matrix. This matrix only holds visitor data by year, not by site. 

```{r}
#combine data
all.years <- rbind(tv97, tv98, tv01, tv12, tv18)
#remove drought column making new dataframe
all.tv <- all.years[, -c(10, 11)]
#create value with drought information
Drought <- c(all.years$Drought)
```


```{r}
# run a NMDS ordination
nmds <- metaMDS(all.tv, distance = "horn")

nmds

plot(nmds, type = "t") 
  ordiellipse(nmds, Drought, col=1:4)

ordiplot(nmds,type="n")
orditorp(nmds,display="species",col="black",air=0.01, cex = .5)
orditorp(nmds,display="sites", col=c(rep("darkgreen", 9), rep("firebrick", 33)),cex=.5,air=0.01)
ordiellipse(nmds, Drought, draw = "polygon", col=c("darkseagreen1", "darksalmon")) 

```

Summary:

The NMDS plot shown in this figure visualizes the difference in space between drought and non-drought years. The two areas show some overlap intidating partial similarities in composition. Some sites fall away from the expexted drought and non-drought years, indiciating within year differences. 


Becasue stress = 0.18, goodness of fit is fair, indicating some misleading distances.
```{r}
data.envfit <- envfit(nmds,Drought)
plot(data.envfit)
data.envfit
```

